peters_elements_causal_inference

### 2021-04-22 paper/draft.Rmd
essay

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: "April 2021"
output: 
  pdf_document: 
    number_sections: yes
    fig_caption: yes
citation_package: --biblatex
fontsize: 12
linestretch: 1.5
abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models. 
---

@@

In causal research, we are interested in the relationship of our variables outside of a fixed time frame. 
We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods^[other datasets].
The causal model entails a probabilistic specification but also entails additional information.
This additional information is a specification for the model when intervention, outcome or feature distributions are modified (Peters et al.,2017).

@@

## Baysian Causal Models

@@

The most common causal graph is the acyclic directed graph. 
A directed graph is a graph where every edge has a direction.
An un-directed graph has edges but there is no arrow which means that the relationship is unclear.
An acyclic graph is a graph without cycles.
That means that a variable does not cause itself, directly or indirectly.
Given the way we look at time where the relationship between variables is indifferent to time, this representation provides a foundation for graphical models.

@@

>>>>

---
title: "A Gentle Introduction into Structural Causal Models"
author: 
- Daniel Saggau $\boldsymbol{\cdot}$ `daniel.saggau@campus.lmu.de`
- Department of Statistics, Ludwig Maximilian University Munich, Germany
date: "April 2021"
output: 
  pdf_document:

    number_sections: yes
    fig_caption: yes
citation_package: --biblatex
fontsize: 12
linestretch: 1.5
header-includes:
- \usepackage{amsmath}
- \usepackage[makeroom]{cancel}
abstract: This paper provides a gentle introduction into causal inference focusing on structural causal models. 
---

@@

In causal research, we are interested in the relationship of our variables outside of a fixed time frame. 
We want to capture relationships that hold outside of an available dataset and results are intended to be generalizable to other time periods^[other datasets].
The causal model entails a probabilistic specification but also entails additional information.
This additional information is a specification for the model when intervention, outcome or feature distributions are modified (Peters et al.,2017).
Probability theory deals with outcomes given a specific underlying data generating process.

@@

## Bayesian Causal Models

@@

The most common causal graph is the acyclic directed graph. 
A directed graph is a graph where every edge has a direction.
An un-directed graph has edges but there is no arrow which means that the relationship is unclear.
We can depict confounding variables via a edge with two arrows.
An acyclic graph is a graph without cycles.
That means that a variable does not cause itself, directly or indirectly.
Given the way we look at time where the relationship between variables is indifferent to time, this representation provides a foundation for graphical models.

@@

## Reichenbach common cause principle

If two [random] variables X and Y are dependent when ignoring z and are independent when accounting for z is the true underlying causal influence.
We can observe Z by looking at the model independent of Z ($X\bcancel{\perp\!\!\!\perp}Y$) and once dependent on Z ($X\perp\!\!\!\perp Y|Z$).(Peters et al., 2017)
Note that there may be a different explanation for the dependence (selection bias)

## Independence Assumption

### 2021-04-22 references/peters_Causality.Rmd
draft1
essay

Probability theory: 
Statistical learning:

>>>>

Probability theory:

Statistical learning: Capacity measures

Peters et al. (2017) suggest that statistical learning theory has a inverse problem: We have data and want to understand the underlying data generating process and respective underlying structures. 
This true underlying data generating process is unknown.
We cannot observe the true DGP based on observational data, alone.
Learning theory proposes the usage of functional specifications of our model to define the true underlying process that is rule based.
These rules are pivotal because they allow us to specify the behavior of the model when exposed to unseen observation data.
To ensure that this specification works, Peters et al. (2017) suggests that statistical learning theory uses the concept of capacity measures. 
If we have high capacity, our functional specification is very applicable to different dataset.
If we have low capacity, our specification is not representative of many datasets.

Apart from having the problem of statistical ill posedness (once because the conditional expectation is undefined for values outside of our sample). 
The second problem is that we usually have no full information for a underlying causal model even when having a fully specified observational distribution.
